SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-10-19 paper_1.tex
Latex source update

In this paper, we introduce perceptron learning to help make this prefetching decision efficiently. We've taken a state of the art prefetcher - Signature Path Prefetcher (SPP) as the underlying prefetch engine. The perceptron layer filters out the unnecessary prefetches recommended by SPP. We've also explored a range of features that can be used to train the perceptron layer. Our results show that SPP-Perc improves performance on SPEC 2017 Benchmark suite by 3.2\% on single-core and 5.5\% on 4-core simulation, as compared to just SPP.

@@

An ideal prefetching scheme involves capturing the memory access patterns to predict future prefetches in a timely manner. Other important decisions are when to bring in the prefetch block of data, which level of cache hierarchy to place the block in and finally, what block to evict to accommodate the incoming block. Memory accesses patterns can be as simple as fetching the next cache line or a complex but predictable pattern, like in pointer chasing applications. Predicting this often requires a trade-off between coverage and accuracy. For a prefetcher to capture highly irregular access patterns, it needs to be highly aggressive (high coverage). That also means the prefetcher would end up recommending a lot of useless prefetches, leading to cache pollution (low accuracy). This effect would get even more pronounced in multi-core scenario where the last level cache is a shared resource. Conversely, a conservative prefetcher would be highly accurate but might not make enough useful predictions.

To this effect we propose <come up with a catchy name> <SPP-Perc>. It's an enhancement to the existing state of the art prefetcher and overcomes the coverage vs trade-off issue in a robust manner. The idea is to modify SPP design to make it as aggressive as it can get. This enables SPP to capture complex memory access patterns hence leading to increased coverage. The prefetch suggestions recommended by SPP pass through the perceptron based throttler. Over time, perceptron layer learns to correlate the prefetches recommended by SPP with various available features and eliminate useless prefetches. This leads to an overall increased accuracy of the prefetch scheme and reduction in DRAM traffic.

@@

\item Designing the prefetcher keeping SPEC 2017 benchmark suite in mind. We've used SPEC 2017 to compare SPP-Perc against the highest performing prefetchers. We've also used our prefetcher to characterize memory intensive benchmarks of the SPEC 2017 suite, with respect to prefetching.

@@

The work was further extended by Multiperspective Reuse Prediction[cite], which introduced a varied set of parametric features. Although design space exploration for feature and parameter selection was a non-trivial task, once fixed, it could give a cache replacement predictor highly suitable for the given set of applications.
\newline
\newline
\textit{Deviations From Actual Perceptrons:} Traditionally, a perceptron prediction involves multiplying the vector of input features: F\textsubscript{1xN}, with the corresponding weight vector: W\textsubscript{Nx1} in a dot product fashion to obtain the sum y\textsubscript{out}. Here what we use is a perceptron-like structure. The feature is used to hash into the weight of perceptrons and the retrieved weights are added straight away. This can be seen a dot product of $\vec{1}$\textsubscript{1xN} with W\textsubscript{Nx1}. This way, the perceptron algorithm doesn't introduce any multiplication operations in the inference or training process. Hence what we adapt in this work, is a perceptron-like learning algorithm as it involves the same principle involved in perceptron inferencing and training. For simplicity, we'll call our implementation as perceptron implementation.
\newline
\newline
\textit{Perceptron Update Rule:} In all the above implementations, a uniform perceptron update principle is being followed. The weights need to be updated if the prediction was wrong or the predicted sum does not cross a certain threshold. After a certain training period, these weights are proportional to the probability of the outcome being true. This training threshold makes sure that the perceptrons are trained till a certain level of confidence and yet they are not over-trained to the given set. Care is taken that the weights saturate at certain positive and negative values so that they remain confined to the bit-width. The same rules will be applied in the discussion done in section 3.2 on SPP-Perc learning algorithm.

\subsection{Signature Path Prefetcher Architecture}
Since the work we've done builds heavily on SPP[cite], we're presenting a brief recap of the ideas introduced in the original paper, especially with an eye towards their relevance in our work. Signature Path Prefetcher (SPP) is a lookahead prefetcher and consists of following structural components:
\newline
\newline
\textbf{Signature Table:} ST keeps track of 256 most recently accessed pages. It is meant to capture memory access patterns within a page boundary. SPP indexes into an entry of ST using the page number. For each entry corresponding to a page, ST stores the 'last block offset' and 'old signature'. Last block offset is the block offset of the last memory access of that given page. The block offset is calculated with respect to the page boundary. The signature is a 12-bit compressed representation of the past few memory accesses for that page. The signature is calculated as:
$$New Signature = (\,Old Signature << 3 bits\,) \;\;XOR\;\; (\,Delta\,)$$ 
In case a matching page entry is found, the stored signature in the corresponding entry is retrieved.
\newline
\newline
Consider the case that incoming page number 10 with a block offset 3 finds a match in the ST. The retrieved pattern signature is 0x30 and the Last Offset is 1. Since now there is the Last Block Offset (1), Incoming Block Offset (3), Old Signature (0x30), and New Signature calculated as per above equation (0x182), SPP can infer non-speculatively that the given pattern of memory accesses (as captured in the Old Signature) leads to the particular Delta. In general, Delta is defined as the difference between the prefetch suggestion block and the initial block which triggered the prefetch. In this case, since SPP is in learning phase, it is defined as the difference between the Incoming Block Offset and the Last Block Offset (+2 in this case). This, in turn generates the new memory pattern (New Signature). This newly learned signature pattern and the delta is stored in the Pattern Table.
\newline
\newline
\textbf{Pattern Table:} The PT is indexed by the signature generated from ST. PT holds predicted delta patterns and their confidence estimates. Each entry indexed by the signature holds up to 4 unique delta predictions. This is implemented by making PT as a 4-way associative table. To measure an estimate of a delta confidence, PT implements two counters: C\textsubscript{sig} for each entry indexed by the signature and C\textsubscript{delta} for each delta entry for a given signature. More details of this on-line score-keeping mechanism is described under Confidence Tracking discussion.

@@

\item \textbf{PF\_THRESHOLD reduced}\newline
SPP design used this threshold to decide whether to do prefetch or not. This threshold (T\textsubscript{p}) was set to 25. Since the new perceptron mechanism is used to make this decision now, this was reduced all the way down to 1. It was not set to 0 for preventing non-throttled lookahead running. SPP confidence mechanism multiplies the confidence of each lookahead depth with the previous one to get the cumulative confidence - on the integer scale of 0 to 100. It was observed that once the lookahead depth falls below one, it essentially becomes 0 since we're limiting the multiplication to integer only. Confidence multiplication for subsequent lookahead depths still yield 0 and prefetching keeps happening. This causes a lot of unnecessary prefetches, leading to cache pollution. To mitigate this, the PF\_THRESHOLD was set to the next possible value i.e. 1.

@@

The neural network weights are stored in a 12-bit indexed (4096 entries deep) table. A single entry in the table is a 5 bit counter - saturating at -16 or +15. Each feature gets its own dedicated table. Our proposed design uses 17 features. The design space exploration and the intiotion behind these features is explained in section\_\_. Hence there are 4096 * 17 5-bit entries which hold the weights of the perceptron layer. As the program execution begins, all the weights are initialized to 0.
\newline
\newline
\textbf{INFERENCING}\newline
In the Champsim simulator, the prefetching method is triggered at L2 Cache demand access. At that point, the prefetcher has the option to do or not do prefetch - in case it does, then how many cache lines to prefetch. The suggested prefetches can be places either in L2 Cache or in Last Level Cache. The prefetcher gets to decide this, and is usually done based on an internal confidence mechanism.

@@

If a demand access triggers the training and the block was tagged as a valid prefetch in the Prefetch Filter, it means that the earlier prefetch prediction was correct. In that case the perceptron weights are incremented by 1 if the predicted sum does not cross a pre-defined threshold.  In case it is a cache block eviction that led to training and the corresponding valid entry was found in the Prefetch Filter, it shows that the prediction made by the perceptron was wrong. Perceptron should have ideally rejected the prefetch suggestion as a low-confidence prefetch. Here the weights are decremented by 1 to reflect the same. In either case, weights are saturated at -16 or +15.

>>>>

In this paper, we introduce perceptron learning to help make this prefetching decision efficiently. We have taken a state of the art prefetcher - Signature Path Prefetcher (SPP) as the underlying prefetch engine. The perceptron layer filters out the unnecessary prefetches recommended by SPP. We have also explored a range of features that can be used to train the perceptron layer. Our results show that SPP-Perc improves performance on SPEC 2017 Benchmark suite by 3.2\% on single-core and 5.5\% on 4-core simulation, as compared to just SPP.

@@

An ideal prefetching scheme involves capturing the memory access patterns to predict future prefetches in a timely manner. Other important decisions faced by the prefetcher are when to bring in the prefetch block of data, which level of cache hierarchy to place the block in and finally, what block to evict to accommodate the incoming block. Memory accesses patterns can be as simple as fetching the next cache line or a complex but predictable pattern, like in pointer chasing applications. Predicting this often requires a trade-off between coverage and accuracy. For a prefetcher to capture highly irregular access patterns, it needs to be highly aggressive (high coverage). That also means the prefetcher would end up recommending a lot of useless prefetches, leading to cache pollution (low accuracy). This effect would get even more pronounced in multi-core scenario where the last level cache is a shared resource. Conversely, a conservative prefetcher would be highly accurate but might not make enough useful predictions.

To this effect we propose <come up with a catchy name> <SPP-Perc>. It's an enhancement to the existing state of the art prefetcher and overcomes the coverage vs accuracy trade-off in a robust manner. The idea is to modify SPP design to make it as aggressive as it can get. This enables SPP to capture complex memory access patterns hence leading to increased coverage. The prefetch suggestions recommended by SPP pass through the perceptron based throttler. Over time, perceptron layer learns to correlate the prefetches recommended by SPP with various available features and eliminate useless prefetches. This leads to an overall increased accuracy of the prefetch scheme and reduction in DRAM traffic.

@@

\item Designing the prefetcher keeping SPEC 2017 benchmark suite in mind. We have used SPEC 2017 to compare SPP-Perc against the highest performing prefetchers. We have also used our prefetcher to characterize memory intensive benchmarks of the SPEC 2017 suite, with respect to prefetching.

@@

The work was further extended by Multiperspective Reuse Prediction[cite], which introduced a varied set of parametric features. Although design space exploration for feature and parameter selection was a non-trivial task, once fixed, it could give a cache replacement predictor highly suitable for the given set of applications.
\newline
\newline
\textit{Deviations From Actual Perceptrons:} Traditionally, a perceptron prediction involves multiplying the vector of input features: F\textsubscript{1xN}, with the corresponding weight vector: W\textsubscript{Nx1} in a dot product fashion to obtain the sum y\textsubscript{out}. Here what we use is a perceptron-like structure. The feature is used to hash into the weight of perceptrons and the retrieved weights are added straight away. This can be seen a dot product of $\vec{1}$\textsubscript{1xN} with W\textsubscript{Nx1}. This way, the perceptron algorithm doesn't introduce any multiplication operations in the inference or training process. Hence what we adapt in this work, is a perceptron-like learning algorithm as it involves the same principle involved in perceptron inferencing and training. For simplicity, we will call our implementation as perceptron implementation.
\newline
\newline
\textit{Perceptron Update Rule:} In all the above implementations, a uniform perceptron update principle is being followed. The weights need to be updated if the prediction was wrong or the predicted sum does not cross a certain threshold. After a certain training period, these weights are proportional to the probability of the outcome being true. This training threshold makes sure that the perceptrons are trained till a certain level of confidence and yet they are not over-trained to the given set. Care is taken that the weights saturate at certain positive and negative values so that they remain confined to the bit-width. The same rules will be applied in the discussion done in section 3.2 on SPP-Perc learning algorithm.

\subsection{Signature Path Prefetcher Architecture}
Since the work we have done builds heavily on SPP[cite], we are presenting a brief recap of the ideas introduced in the original paper, especially with an eye towards their relevance in our work. Signature Path Prefetcher (SPP) is a lookahead prefetcher and consists of following structural components:
\newline
\newline
\textbf{Signature Table:} ST keeps track of 256 most recently accessed pages. It is meant to capture memory access patterns within a page boundary. SPP indexes into an entry of ST using the page number. For each entry corresponding to a page, ST stores the 'last block offset' and 'old signature'. Last block offset is the block offset of the last memory access of that given page. The block offset is calculated with respect to the page boundary. The signature is a 12-bit compressed representation of the past few memory accesses for that page. The signature is calculated as:
$$New Signature = (\,Old Signature << 3 bits\,) \;\;XOR\;\; (\,Delta\,)$$ 
In case a matching page entry is found, the stored signature in the corresponding entry is retrieved.
\newline
\newline
Consider the case that incoming page number 10 with a block offset 3 finds a match in the ST. The retrieved pattern signature is 0x30 and the Last Offset is 1. Since now there is the Last Block Offset (1), Incoming Block Offset (3), Old Signature (0x30), and New Signature calculated as per above equation (0x182), SPP can infer non-speculatively that the given pattern of memory accesses (as captured in the Old Signature) leads to the particular Delta. In general, Delta is defined as the difference between the prefetch suggestion block and the initial block which triggered the prefetch. In this case, since SPP is in learning phase, it is defined as the difference between the Incoming Block Offset and the Last Block Offset (+2 in this case). This, in turn generates the new memory pattern (New Signature). This newly learned signature pattern and the delta is stored in the Pattern Table.
\newline
\newline
\textbf{Pattern Table:} The PT is indexed by the signature generated from ST. PT holds predicted delta patterns and their confidence estimates. Each entry indexed by the signature holds up to 4 unique delta predictions. This is implemented by making PT as a 4-way associative table. To measure an estimate of a delta confidence, PT implements two counters: C\textsubscript{sig} for each entry indexed by the signature and C\textsubscript{delta} for each delta entry for a given signature. More details of this on-line score-keeping mechanism is described under Confidence Tracking discussion.

@@

\item \textbf{PF\_THRESHOLD reduced}\newline
SPP design used this threshold to decide whether to do prefetch or not. This threshold (T\textsubscript{p}) was set to 25. Since the new perceptron mechanism is used to make this decision now, this was reduced all the way down to 1. It was not set to 0 for preventing non-throttled lookahead running. SPP confidence mechanism multiplies the confidence of each lookahead depth with the previous one to get the cumulative confidence - on the integer scale of 0 to 100. It was observed that once the lookahead depth falls below one, it essentially becomes 0 since we are limiting the multiplication to integer only. Confidence multiplication for subsequent lookahead depths still yield 0 and prefetching keeps happening. This causes a lot of unnecessary prefetches, leading to cache pollution. To mitigate this, the PF\_THRESHOLD was set to the next possible value i.e. 1.

@@

The neural network weights are stored in a 12-bit indexed (4096 entries deep) table. A single entry in the table is a 5 bit counter - saturating at -16 or +15. Each feature gets its own dedicated table. Our proposed design uses 17 features. The design space exploration and the intuition behind these features is explained in section\_\_. Hence there are 4096 * 17 5-bit entries which hold the weights of the perceptron layer. As the program execution begins, all the weights are initialized to 0.
\newline
\newline
\textbf{INFERENCING}\newline
In the Champsim simulator, the prefetching method is triggered at L2 Cache demand access. At that point, the prefetcher has the option to do or not do prefetch - in case it does, then how many cache lines to prefetch. The suggested prefetches can be places either in L2 Cache or in Last Level Cache. The prefetcher gets to decide this, and is usually done based on an internal confidence mechanism.

@@

If a demand access block which triggers the training  was tagged as a valid prefetch in the Prefetch Filter, it means that the earlier prefetch prediction was correct. In that case the perceptron weights are incremented by 1 if the predicted sum does not cross a pre-defined threshold.  In case it is a cache block eviction that led to training and the corresponding valid entry was found in the Prefetch Filter, it shows that the prediction made by the perceptron was wrong. Perceptron should have ideally rejected the prefetch suggestion as a low-confidence prefetch. Here the weights are decremented by 1 to reflect the same. In either case, weights are saturated at -16 or +15.
